---
title: "NOAA Storm Database - Analysis: Severe weather events"
author: "Lucas Carpio"
date: "03 January 2021"
output:
        html_document:
                keep_md: true
---

# Synopsis

Analyse the impact that different type of Climate events have in the health and economics
in the USA. Data from NOAA storm data base is used for the analysis. 

For the Health analysis, the objective is to identify which events are the most harmful. 
For the economical impact, validate which climate events have caused the most expensive
losses in Properties and in Crops.

```{r setup, include=FALSE}
rm(list = ls()) #Delete existing data in global environment
library(knitr)
library(lattice)
library(ggplot2)
library(dplyr)
library(anytime)
opts_chunk$set(echo = TRUE, cache = TRUE) ## Always show answers and avoid repeating unnecessary calculations
```

# Questions:

The data Analysis will address the following questions:


- Across the United States, which types of events (as indicated in the EVTYPE variable) are
most harmful with respect to population health?


- Across the United States, which types of events have the greatest economic consequences?



# Data processing

## Reading the Data

``` {r readingData}
# for Reading Files *ZIP
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, "dataset.zip")
Rawdata <- read.csv("dataset.zip", header = TRUE, sep = ",", check.names = TRUE, stringsAsFactors=FALSE)
``` 


The variable **BGN_DATE** represent a Date and is being shown as "Factor". To convert to type: "Date"


``` {r changetodate}
#Make a copy of data to avoid reading again the file
Rawdata2 <- Rawdata
Rawdata2$BGN_DATE <- as.Date(Rawdata2$BGN_DATE, "%m/%d/%Y")
#validate if type of data for "Date" has changed
str(Rawdata2$BGN_DATE)
```

```{r changedatetype, include=FALSE}
#Check data
str(Rawdata2)
```



## First Question

**The first Questions is to know which Event Type has been more harmful with respect to
population.**

The first approach is to validate wich Event Type has occurred more times

```{r MostrepeatedEvent}
EventTypeSummary <- table(Rawdata2$EVTYPE)
EventTypeSummary <- data.frame(EventTypeSummary)
EventTypeSummary <- arrange(EventTypeSummary, desc(Freq))
head(EventTypeSummary, 5)
``` 

From this table, we can assure that **`r EventTypeSummary$Var1[[1]]`** is the event that happens
the most in USA. Nevertheless, it does not guarantee that it is the most harmful for the 
population.

To validate the Harmfulness, the **"Fatalities"** and **"Injuries"** caused by the events will be
analyzed.


```{r fatalitiesInjuries}
FatalitiesInjuries <- subset(Rawdata2, select =  c(EVTYPE, FATALITIES, INJURIES))
EvFatalities <- aggregate(FatalitiesInjuries$FATALITIES, by = list(FatalitiesInjuries$EVTYPE), FUN = sum)
EvInjuries <- aggregate(FatalitiesInjuries$INJURIES, by = list(FatalitiesInjuries$EVTYPE), FUN = sum)
#Order the Data from Higher to lower to see the events with more fatal and injure events:

EvFatalities <- arrange(EvFatalities, desc(x))
EvInjuries <- arrange(EvInjuries, desc(x))
HarmfulEvents <- merge(EvFatalities, EvInjuries, by = "Group.1", all = TRUE)
HarmfulEvents <- arrange(HarmfulEvents, desc(x.x))
names <- c("EVTYPE", "FATALITIES", "INJURIES")
colnames(HarmfulEvents) <- names
head(HarmfulEvents, 5)
```

Here, one can see that **TORNADO** and **EXCESSIVE HEAT** have the higher values in Fatalities.
Comparing with the frequency of the events (Table shown above "EventTypeSummary"), the 
Frequency of the events is not directly related with the Harmfulness. 

Considering that the Fatal cases are worst as the Injuries, but without taking out of consideration
the injuries, a Weighted calculation is made, where a weight of "5" is given to "Fatality"
and a Weight of "1" to "Injury"

```{r weightedvalue}
HarmfulEvents2 <- HarmfulEvents
HarmfulEvents2 <- HarmfulEvents2 %>% mutate(WeightedValue = FATALITIES*5 + INJURIES)
HarmfulEvents2 <- arrange(HarmfulEvents2, desc(WeightedValue))
head(HarmfulEvents2, 5)
```

With this conisderation, the most harmful event is the **TORNADO**, followed by the **EXCESIVE HEAT**.


To be clear with the **"Weighted value"**, this result might vary according to the weight given
to the "FATALITY". To validate its impact, another calculation with a higher weight is made to see
if the top 3 events vary:


```{r weightedvalue2}
HarmfulEvents2 <- HarmfulEvents
HarmfulEvents2 <- HarmfulEvents2 %>% mutate(WeightedValue = FATALITIES*10 + INJURIES)
HarmfulEvents2 <- arrange(HarmfulEvents2, desc(WeightedValue))
head(HarmfulEvents2, 3)
```

### Answer Question 1
* It is clear that **TORNADO** and **EXCESIVE HEAT** are the 2 most harmful events in the USA.


## Second Question

**Identify which types of events have the greatest economic consequences**

Identify which variables are related with Economic Consecuences:

* **PROPDMG**   : Property Damage. Number of Property damaged
* **PROPDMGEXP**: Property Damage Exponential. " ",-, ?, +, 0, 1, 2, 3, 4, 5, 6, 7, 8, B, h, H, K, m, M,
* **CROPDMG**   : Crop Damage. Number of Crop damage
* **CROPDMGEXP**: Crop Damage Exponential. ?,  0,  2,  B,  k,  K,  m,  M

According to a validation made by _Soesilo Wijono_ in an article published in RPubs,
[RPubs Link](https://rpubs.com/flyingdisc/PROPDMGEXP), the equivalents are as follows:

* H,h = hundreds = 100

* K,k = kilos = thousands = 1,000

* M,m = millions = 1,000,000

* B,b = billions = 1,000,000,000

* (+) = 1

* (-) = 0

* (?) = 0

* black/empty character = 0

* numeric 0..8 = 10


To avoid working with all complete data set, a new data set is created using only the 
necessary variables

```{r subset2question}
EconomicVariables <- subset(Rawdata2, select =  c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
str(EconomicVariables)
```


Using the previous table, the equivalents values are assigned accordingly for __Property__

```{r assingvaluestoEXPProp}
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "H"] <- 100
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "h"] <- 100
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "K"] <- 1000
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "k"] <- 1000
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "M"] <- 1000000
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "m"] <- 1000000
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "B"] <- 1000000000
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "b"] <- 1000000000
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "+"] <- 1
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "-"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "?"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == " "] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == ""] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "0"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "1"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "2"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "3"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "4"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "5"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "6"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "7"] <- 0
EconomicVariables$PROPEXP[EconomicVariables$PROPDMGEXP == "8"] <- 0
```

Using the previous table, the equivalents values are assigned accordingly for __Crop__

```{r assignvaluestoEXPCrop}
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == ""] <- 0
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == "M"] <- 1000000
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == "K"] <- 1000
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == "m"] <- 1000000
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == "B"] <- 1000000000
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == "?"] <- 0
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == "0"] <- 0
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == "k"] <- 1000
EconomicVariables$CROPEXP[EconomicVariables$CROPDMGEXP == "2"] <- 0
```

Having the equivalences for both _PROPERTY_ and _CROP_, we can calculate the final value

```{r EconomicImpact}
EconomicVariables$PROPValue <- EconomicVariables$PROPDMG * EconomicVariables$PROPEXP
EconomicVariables$CROPValue <- EconomicVariables$CROPDMG * EconomicVariables$CROPEXP

# To see the top 5 Types of events with most Property Damage
top10Prop <- arrange(EconomicVariables, desc(PROPValue))[1:10, ]
head(top10Prop, 5)

# To see the top 5 Types of events with most Crop Damage
top10crop <- arrange(EconomicVariables, desc(CROPValue))[1:10, ]
head(top10crop, 5)
```


### Answer Question 2
* In _Properties_, the biggest consecueces are due to **FLOOD** and for _Crops_ are due to
**River Flood**.


# Results

## Question 1

```{r results1}
par(mfrow = c(1,3), mar = c(12,3,2,2), mgp = c(3,1,0), cex = 0.7)
barplot(HarmfulEvents2$FATALITIES[1:10], las = 3, names.arg = HarmfulEvents2$EVTYPE[1:10], main = "Top 10 Wheater Fatalities", ylab = "# fatalities", col = "yellow")
barplot(HarmfulEvents2$INJURIES[1:10], las = 3, names.arg = HarmfulEvents2$EVTYPE[1:10], main = "Top 10 Wheater Injuries", ylab = "# Injuries", col = "blue")
barplot(HarmfulEvents2$WeightedValue[1:10], las = 3, names.arg = HarmfulEvents2$EVTYPE[1:10], main = "Top 10 Harm Events - Weighted", ylab = "# Harm - Weighted", col = "red")
```

**Tornado** is the event that causes more harm in USA, Followed by **Excessive Heat**. 
The third graph is a weighted representation of the events that causes more harm,
considering fatality events worse than injuries (Weighted calculation).

## Question 2
```{r results2}
par(mfrow = c(1,2), mar = c(12,3,2,2), mgp = c(3,1,0), cex = 0.7)
barplot(top10Prop$PROPValue, las = 3, names.arg = top10Prop$EVTYPE, main = "Top 10 Property Damage", ylab = "$ Property loss", col = "cyan")
barplot(top10crop$CROPValue, las = 3, names.arg = top10crop$EVTYPE, main = "Top 10 Crop Damage", ylab = "$ Crop loss", col = "green")
```

For **Property** the biggest loosses are casued by **Floods**.
For **Crop** the biggest losses are caused by **River Floods** 


.
.


Complete project and information available in [GitHub Link](https://github.com/luquencio/Severe_weather_events)

.
.


End of Document
